Large-scale hybrid ad hoc network for mobile platforms: Challenges and Experiences

### 2019-02-19 paper.tex
Final submitted version

\section{Introduction}
Mobile devices with programmable platforms such as android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark \footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been widely studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on analytical and simulation-based study of
MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges in deploying a large number of
physical nodes, real-world implementations have been limited and have
not reached mass scale \cite{kiess-manet-impl-2007}. However, with the
growth of smartphones, large-scale real-world implementations may
become feasible. This paper describes a real-world implementation of a
p2p delay-tolerant network, called Mesh Network Alerts (MNA), for
relaying severe weather information to millions of mobile device users
as part of the Weather Channel app \cite{mna} on both android and iOS
platforms.

Before describing MNA, it is critical to identify applications that
need p2p communication, given pervasive Internet connectivity. Doing
so enables us to define key characteristics of such applications and
focus on the challenges in meeting them.  This paper focuses on two
separate classes of applications: communication in
(a) disaster-affected or remote areas and (b) congested networks in
densely populated areas, e.g., sports arenas.  The following are the
key characteristics in these scenarios:

%
\btask
%
\item\label{c:0} No communication infrastructure such as WiFi
  access points to fall back on
\item\label{c:1} Network nodes are mobile, pattern of mobility is not
  predictable
\item\label{c:2} New information may arrive at any time
\item\label{c:3} Trustworthy information is scarce, misinformation and
  rumours are common place
\item\label{c:4}  Small payloads suffice in many cases and information
  retains value for a few minutes
\item\label{c:5}  Device battery is a scarce resource, power supply
  for recharging may not be available
\item\label{c:6}  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
%
\etask
%

@@

A vast majority of the literature has focused on a traditional model
of stateful, fully decentralized, reliable networking. Specifically,
the nodes maintain connectivity with peers and routing is optimized
with techniques based on link state or distance vectors
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization.

Given the application characteristics above, this paper identifies
main practical challenges associated with modern device platforms and
finds novel ways to overcome them.  This leads to MNA --- a new
paradigm in p2p networking that employs a hybrid, delay-tolerant, and
zero-routing overhead architecture.  Unlike previous MANET
architectures, MNA is a hybrid of a centralized and a decentralized
architecture, called \emph{center-to-peer-to-peer (c2p2p)}. A central service
is leveraged as the trusted source of information while the
information is propagated in a decentralized fashion by peers. Such an
architecture allows MNA to bypass the key issue of trust deficit in
open decentralized systems and scale while retaining the ability of
infrastructure-less communication in many application scenarios.

@@

This paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel c2p2p architecture implementation using multiple radio
  channels for modern mobile platforms (Android and iOS)
\item Experimental evaluation and deployment statistics
\end{itemize}

@@

Android and iOS in addressing the challenges. Experimental evaluation
and deployment statistics are presented in Section~\ref{sec:eval}.  A
deeper look at the literature and contrast to MNA is summarized in
Section~\ref{sec:related} with conclusions in

@@

continue to discover peer devices to receive and forward

@@

and iOS offer APIs to discover and advertise information to peer
devices over WiFi and Bluetooth interfaces, peer-to-peer connections
do not work reliably when the same APIs are accessed while the app is
in the background. Further, on Android, each WiFi p2p connection must
be explicitly approved by the device user. Prior works widely document

@@

MNA functions as expected on a single device. Further, running test

@@

large number of devices need to take specific coordinated action

@@

model and radio.  This is not a challenge in traditional mobile

@@

As user devices with MNA advertise on unsecured wireless protocols, it
may be possible for a malicious attacker to listen for such
advertisements and reverse-engineer the message formats and protocols
used. Then, the attackers may generate fake messages and advertise
them, e.g., a fake tornado alert. Given a lack of trusted information
in such scenarios (\ref{c:3}), misinformation campaigns can have
disastrous consequences.  In open decentralized systems, such false
messages cannot be distinguished from the real ones, and MNA will end
up propagating them to as many devices as possible, ``poisoning'' the

@@

independently verified in open decentralized systems and previous works
on peer-to-peer networks do not address this challenge.

@@

To address (or bypass) the challenges identified above, this paper
proposes c2p2p -- a hybrid architecture that leverages a central
service as the sole source of trusted information as depicted in
Figure~\ref{fig:arch}. An application-specific central service, e.g.,
weather.com backend, originates new information as a push message and
assigns it a unique identifier. Next, the central service produces a
digital signature using its private key and appends it to the message
payload. Finally, header parameters specifying TTL (duration after
which the message expires), peer identifier of the central service,
destination peer identifier list (for uni-cast) or a special identifier
(for broadcast), and the time of origin are added to the message and
the message is sent to a push notification service for distribution.

Mobile applications integrate MNA as an SDK including the public key
of the corresponding central service. Peers having Internet
connectivity receive the push messages and verify the digital
signature to protect against spread of misinformation.  If the message
has not expired and is destined to other peers, then the receiving
peers start forwarding the message to other peers. Depending on the
protocol being used, such forwarding happens as a uni-cast or
broadcast.

Since new information can arrive at any time or new peers may come
into proximity, and message transmissions are not reliable, all peers
store unexpired and verified messages and continue to forward them to
proximal peers repeatedly. This implies that peers may receive the
same message more than once but since the messages are globally
unique, duplicates can be identified and ignored.  A peer sending a
message adds its own peer identifier to a list of forwarders appended
to the message header.  A receiving peer thus knows all the peers that
already have the message and stops repeating the message to those
peers, controlling the flooding.  As there are no control messages or
any other overhead for routing, we call this \emph{zero-overhead}
routing.

The following assesses the c2p2p architecture relative to the
challenges identified above along with known limitations. Further, to
realize this architecture on mobile devices, we evaluated a variety of
protocols on both Android and iOS. As described later, WiFi DNS on
Android, Bluetooth Nearby on Android, and Bluetooth low energy (BTLE)
on both Android and iOS are the only protocols that were found
effective. Actual techniques for discovery, advertisement, and
connection vary across these protocols and are described next.
%
\subsection{Operating system restrictions}
\label{sec:os}
MNA works around the background activity and user permission
restrictions via innovative techniques, without resorting to hacks
that may violate user security or App-store guidelines.
Tables~\ref{tab:android} and~\ref{tab:ios} summarize the various
protocols we have experimented with using WiFi and Bluetooth
interfaces, along with key findings as described here.  The protocols
actually deployed are in bold.
%
%
\begin{table}[H]
\caption{Android protocols and key findings}
\label{tab:android}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{iOS} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{WiFi}}} & \textbf{DNS} & 0.2 & 2\% & 600 & No & \\
& WiDi & 2000 & 3\% & 600 & No & Needs DNS\\
& Hotspot & 2000 & 5\% & 600 & No & Permission\\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{BT}}} & \textbf{Classic} & 50 & 2\% & 600 & No & \\
& \textbf{Nearby} & 50 & 2\% & 600 & No & \\
& \textbf{BTLE} & 50 & 2\% & 600 & Yes & \\
\hline
\end{tabular}
\end{table}
%
%
\begin{table}[H]
\caption{iOS protocols and key findings}
\label{tab:ios}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{Android} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{2}{*}{\rotatebox[origin=c]{90}{WiFi}}} & Bonjour & 0.2 & 5\% & 200 & No & Battery\\
& MPC & 2000 & 5\% & 200 & No & Battery\\
\hline
& \textbf{BTLE} & 50 & 1\% & 800 & Yes & \\
\hline
\end{tabular}
\end{table}
%

Line of sight range in feet was measured by manually placing devices
at increasing distances until they fail to communicate. Battery
consumption per hour was measured as a difference between baseline
battery consumption and when MNA was left running for 12
hours. Reported numbers are an average percentage consumption across
about 50 devices on each platform.

BTLE is the only cross-platform protocol, although there are
limitations with operating system versions as shown in
Table~\ref{tab:android_ios}. Android is denoted as A, 18 and 21 are
correspond to Android versions 4.3 and 5.0 respectively, and C and P
denote the roles of Central and Peripheral. Key finding is that
Android 4.3 devices acting as Peripheral cannot communicate either as
senders or receivers.
%
\begin{table}[H]
\caption{BTLE interoperability between Android and iOS devices}
\label{tab:android_ios}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline 
\multicolumn{1}{|c|}{Sender} & \multicolumn{6}{c|}{Receiver} \\
\hline
\multicolumn{1}{|c|}{} & A.18-C & A.18-P & A.21-C & A.21-P  & iOS-C & iOS-P \\
\hline
A.18-C & \cellcolor{Gray}         &   No       & \cellcolor{Gray}         &    Yes       &  \cellcolor{Gray}       &   Yes      \\ 
\hline
A.18-P &  No        & \cellcolor{Gray}         &   No       &  \cellcolor{Gray}         &   No      &  \cellcolor{Gray}       \\ 
\hline
A.21-C & \cellcolor{Gray}         &   No       &  \cellcolor{Gray}        &    Yes       &  \cellcolor{Gray}       &   Yes      \\ 
\hline
A.21-P &  Yes        & \cellcolor{Gray}         &   Yes       &  \cellcolor{Gray}         &   Yes      & \cellcolor{Gray}        \\ 
\hline
iOS-C & \cellcolor{Gray}         &    No      &  \cellcolor{Gray}        &     Yes       & \cellcolor{Gray}        &   Yes      \\ 
\hline
iOS-P &   Yes       & \cellcolor{Gray}         &    Yes      &  \cellcolor{Gray}         &   Yes      &  \cellcolor{Gray}       \\ 
\hline
\end{tabular}
\end{table}
%

@@

%
\subsection{Power constraints}
\label{sec:power}
%
Due to indefinite foreground services and continuous discovery and
advertisement, this is primarily an issue on Android. Our approach
here is two-pronged. Firstly, via extensive experiments on a large
number of devices, we fine-tune the algorithms governing the intervals
at which discovery and advertisements occur. In a nutshell, receiving
new information during a period causes MNA to be more aggressive in
discovery and advertisement. Similarly, lack of new information for a
period makes the device less aggressive. Secondly, we allow the
central service to broadcast``wake up'' messages ahead of an
anticipated severe weather event.  When devices receive such messages,
they schedule themselves to remain aggressive during the specified
window of time.  Outside of this window, the device can afford to have
long sleep cycles and conserve power. With these techniques, our
testing shows less than 2\% battery consumption per hour on most
device models.
%
\subsection{Testing p2p networks}
\label{sec:texting}
%
We developed novel test automation tools and processes that control
multiple devices from a single test station and follow prescribed
steps to generate, send, and receives messages to play out a test
scenario.  The framework allows automated analysis of the observations
to determine the test result. This capability was instrumental in
uncovering bugs at a fast pace with tens of physical devices employed
for automated testing.  Further, the automation framework is general
and can be expressly applied to test other apps in this fashion.

Figure~\ref{fig:test_arch} shows the overall test automation
architecture. A tester specifies high-level test cases and expected
results in a platform and device agnostic fashion. Mobile devices run
a UI proxy for executing platform and device-specific UI command while
a central laptop terminal acts as the overall orchestrator and
translates high-level test cases and distributes them to
devices. Since the orchestrator needs a wired access to the devices
and the devices cannot be programmed to move (unlike drones and bots),
testing of range and mobility patterns needs to done manually, which
is a known limitation.

A key innovation here is the platform-independent language for
specifying high-level test cases and a small set of primitive commands
it translates to. We describe here the list of primitive commands and
a screen recording across seven Android devices demonstrates this
powerful capability \cite{mna-test-demo}.
\begin{itemize}
\item \textsf{syncAll} Wait for all devices to reach this point
\item \textsf{object.find} Find a UI object by name or position
\item \textsf{object.click} Click a UI object
\item \textsf{sleep} Wait for the specified duration
\item \textsf{setupSystemAlert} Respond to a system popup
\item \textsf{text.set} Highlight a text input box
\item \textsf{text.write} Write text to the text box
\item \textsf{swipe} Swipe to the left/right or up/down
\end{itemize}

@@

%
\subsection{Trust in information}
\label{sec:trust}
%
In c2p2p, since the origin of information is always the central
service, digital signatures can be attached to all messages
originating from the service. The mobile application is distributed
with the corresponding public key so that digital signatures from the
service can be verified.  If a message fails such a verification, it
is discarded and not forwarded any further.

Peers cannot originate messages however as then a malicious peer can
join the network, exchange keys with other peers, start generating,
signing, and spreading fake information. This has been a longstanding
problem for open decentralized systems. Requiring all peers to
register their identity with a certificate authority and reporting the
originators of fake information may discourage malicious behavior.
Clearly, this is even harder to solve than the spread of fake
information over the Internet because ground truth is hard to find and
malicious peers can report genuine sources of information as well.
%
\section{Algorithms}
\label{sec:algo}
%
\subsection{WiFi DNS}
\label{sec:wifi}
%
DNS service discovery is a widely supported protocol
\cite{cheshire-dns-sd-2013}. On Android, WiFi DNS improvises on this
standard such that the exchange of information happens without making
network connections at all. This is achieved by splitting the message
payloads into small enough chunks and stuffing them into service
advertisements as txt records and broadcast over multiple
advertisements in a quick succession. Since there are no connections
being made, the messages can be exchanged without prompting the user
to grant permissions for each connection. Since no peer connections
are needed and advertisements and discovery are broadcast, WiFi DNS is
a broadcast protocol. On iOS, Bonjour protocol is built on DNS service
discovery but is power-hungry and does not inter-operate with WiFi DNS
on Android. Algorithm~\ref{algo:dns} outlines the peer process for
WiFi DNS. Power conservation via time intervals is achieved by
manipulating $T_{ad}$ and $T_{disc}$ based on traffic.

Messages from multiple applications are forwarded on the same network,
but a peer can validate signatures only for the messages targeting an
app it hosts. Peers simply forward messages not targeting one of their
apps. Key benefit of this design is that the density of peers is
available to all apps, making c2p2p scenarios viable even when an
individual app does not have a sufficient density of users.  However,
this also allows a malicious jammer to generate traffic for a fake
app and force the peers to forward it (as they do not match any of the
apps) until the messages expire.  Although the fake information will
never be delivered to the apps, this issue can affect MNA throughput
for legitimate apps. A possible solution would be to maintain a
central registry of $a_{id}$ and distribute it to all peers across
apps.

>>>>

\section{Introduction}
Mobile devices with programmable platforms such as Android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark\footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been widely studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on analytical and simulation-based study of
MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges in deploying a large number of
physical nodes, real-world implementations have been limited and have
not reached mass scale \cite{kiess-manet-impl-2007}. However, with the
growth of smartphones, large-scale real-world implementations may
become feasible. This paper describes a real-world implementation of a
p2p delay-tolerant network, called Mesh Network Alerts (MNA), for
relaying severe weather information to millions of mobile device users
as part of the Weather Channel app \cite{mna} on both Android and iOS
platforms.

Before describing MNA, it is critical to identify applications that
need p2p communication, given pervasive Internet connectivity and the
ease of using it. Doing so enables us to define key characteristics of
such applications and focus on the challenges in meeting them.  This
paper focuses on two separate classes of applications: communication
in (a) disaster-affected or remote areas and (b) congested networks in
densely populated areas, e.g., sports arenas.  The following are the
key characteristics in these scenarios:

%
\btask
%
\item\label{c:0} No communication infrastructure such as WiFi
  access points to fall back on
\item\label{c:1} Device users are mobile, pattern of mobility is not
  predictable
\item\label{c:2} New information may arrive at any time
\item\label{c:3} Trustworthy information is scarce, misinformation and
  rumours are common place
\item\label{c:4}  Small payloads suffice in many cases and information
  retains value for a few minutes
\item\label{c:5}  Device battery is a scarce resource, power supply
  for recharging may not be available
\item\label{c:6}  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
%
\etask
%

@@

A vast majority of the literature has focused on a traditional model
of stateful, fully decentralized, reliable networking. Specifically,
the nodes maintain connectivity with peers and routing is optimized
with techniques based on link state (OLSR) or distance vectors (AODV)
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization.

Given the application characteristics above, this paper identifies
practical challenges associated with modern device platforms and finds
novel ways to overcome them.  This leads to MNA -- a new paradigm in
p2p networking that employs a delay-tolerant, and zero-routing
overhead architecture.  Unlike previous works, MNA is a hybrid of a
centralized and a decentralized architecture, called
\emph{center-to-peer-to-peer (c2p2p)}. A central service is leveraged
as the trusted source of information while the information is
propagated in a decentralized fashion by peers. Such an architecture
allows MNA to bypass the key issue of trust deficit in open
decentralized systems while retaining the ability of
infrastructure-less communication in many application scenarios.

@@

This paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel c2p2p implementation using multiple radio
  channels on modern mobile platforms (Android and iOS)
\item Experimental evaluation and real-world deployment statistics
\end{itemize}

@@

Android and iOS in addressing the challenges. Section~\ref{sec:algo}
dives into the algorithms of two of the MNA protocols. Experimental
evaluation and deployment statistics are presented in
Section~\ref{sec:eval}.  A deeper look at the literature and contrast
to MNA is summarized in Section~\ref{sec:related} with conclusions in

@@

continue to discover peers to receive and forward

@@

and iOS offer APIs to discover and advertise information to peers over
WiFi and Bluetooth interfaces, peer-to-peer connections do not work
reliably when the same APIs are accessed while the app is in the
background. Further, on Android, each WiFi p2p connection must be
explicitly approved by the user. Prior works widely document

@@

MNA works as expected on a single device. Further, running test

@@

large number of devices need to take coordinated action

@@

hardware.  This is not a challenge in traditional mobile

@@

As user devices advertise on unsecured wireless protocols, it may be
possible for a malicious attacker to listen for such advertisements
and reverse-engineer the protocols used. Then, the attackers may
generate fake messages and advertise them, e.g., a fake tornado
alert. Given a lack of trusted information in such scenarios
(\ref{c:3}), misinformation campaigns can have disastrous
consequences.  In open decentralized systems, such false messages
cannot be distinguished from the real ones, and MNA will end up
propagating them to as many devices as possible, ``poisoning'' the

@@

independently verified in open decentralized systems and previous
works on peer-to-peer networks do not address this challenge.

@@

To address (or bypass) the challenges identified above, this paper
proposes c2p2p -- a hybrid architecture that leverages a central
service as the sole source of trusted information as depicted in
Figure~\ref{fig:arch}. An application-specific central service, e.g.,
weather.com backend, originates new information and
assigns it a unique identifier. Next, the central service produces a
digital signature using its private key and appends it to the message
payload. Finally, header parameters specifying TTL (duration after
which the message expires), peer identifier of the central service,
destination peer identifier list (for uni-cast) or a special identifier
(for broadcast), and the time of origin are added to the message and
the message is sent to a push notification service for distribution.

Mobile applications integrate MNA as an SDK with the public key of the
corresponding central service. Peers having Internet connectivity
receive the push messages and verify the digital signature.  If the
signature is valid, the message has not expired, and is destined to
other peers, then the receiving peers forward the message to other
peers. Depending on the protocol being used, such forwarding happens
as a uni-cast or broadcast. Peers receiving the forwarded messages
also verify the digital signature before forwarding further.

Since new information can arrive at any time, new peers may come into
proximity, or message transmissions fail, all peers store unexpired
and verified messages and forward them to proximal peers
repeatedly. This implies that peers may receive a message more than
once but since the messages are globally unique, duplicates can be
identified and ignored.  A peer sending a message adds its own peer
identifier to a list of forwarders appended to the message header.  A
receiving peer thus knows all the peers that already have the message
and stops repeating the message to those peers, controlling the
flooding.  As there are no control messages or any other overhead for
routing, we call this \emph{zero-overhead} routing.

The following assesses the c2p2p architecture relative to the
challenges identified above along with known limitations. Further, to
realize this architecture on mobile devices, we evaluate a variety of
protocols on both Android and iOS. As described later, WiFi DNS on
Android, Bluetooth Nearby on Android, and Bluetooth low energy (BTLE)
on both Android and iOS are the only protocols that were found
effective. Actual techniques for discovery, advertisement, and
connection vary across these protocols and are described next.
%
\subsection{Operating system restrictions}
\label{sec:os}
%
%
\begin{table}[H]
\caption{Android protocols and key findings}
\label{tab:android}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{iOS} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{WiFi}}} & \textbf{DNS} & 0.2 & 2\% & 600 & No & \\
& WiDi & 2000 & 3\% & 600 & No & Needs DNS\\
& Hotspot & 2000 & 5\% & 600 & No & Permission\\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{BT}}} & \textbf{Classic} & 50 & 2\% & 600 & No & \\
& \textbf{Nearby} & 50 & 2\% & 600 & No & \\
& \textbf{BTLE} & 50 & 2\% & 600 & Yes & \\
\hline
\end{tabular}
\end{table}
%

MNA works around the background activity and user permission
restrictions via innovative techniques, without resorting to hacks
that may violate user security or developer guidelines.
Tables~\ref{tab:android} and~\ref{tab:ios} summarize the various
protocols we have experimented with using WiFi and Bluetooth
interfaces, along with key findings as described here.  The protocols
actually deployed are in bold.

%
\begin{table}[H]
\caption{iOS protocols and key findings}
\label{tab:ios}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{Android} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{2}{*}{\rotatebox[origin=c]{90}{WiFi}}} & Bonjour & 0.2 & 5\% & 200 & No & Battery\\
& MPC & 2000 & 5\% & 200 & No & Battery\\
\hline
& \textbf{BTLE} & 50 & 1\% & 800 & Yes & \\
\hline
\end{tabular}
\end{table}
%

Throughput is measured over a single p2p hop. Line of Sight (LoS)
range in feet was measured by manually placing devices at increasing
distances until they fail to communicate. Battery consumption per hour
was measured as a difference between baseline battery consumption and
when MNA was left running for 12 hours. Reported numbers are an
average percentage consumption across about 50 devices on each
platform.

BTLE is the only cross-platform protocol, although there are
limitations with operating system versions as shown in
Table~\ref{tab:android_ios}. Android is denoted as A, 18 and 21 API
levels corresponding to Android versions 4.3 and 5.0 respectively, and
C and P denote the roles of Central and Peripheral. Key finding is
that Android 4.3 devices acting as Peripheral cannot communicate
either as senders or receivers.
%
\begin{table}[H]
\caption{BTLE interoperability between Android and iOS devices}
\label{tab:android_ios}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline 
\multicolumn{1}{|c|}{Sender} & \multicolumn{6}{c|}{Receiver} \\
\hline
\multicolumn{1}{|c|}{} & A.18-C & A.18-P & A.21-C & A.21-P  & iOS-C & iOS-P \\
\hline
A.18-C & \cellcolor{Gray}         &   No       & \cellcolor{Gray}         &    Yes       &  \cellcolor{Gray}       &   Yes      \\ 
\hline
A.18-P &  No        & \cellcolor{Gray}         &   No       &  \cellcolor{Gray}         &   No      &  \cellcolor{Gray}       \\ 
\hline
A.21-C & \cellcolor{Gray}         &   No       &  \cellcolor{Gray}        &    Yes       &  \cellcolor{Gray}       &   Yes      \\ 
\hline
A.21-P &  Yes        & \cellcolor{Gray}         &   Yes       &  \cellcolor{Gray}         &   Yes      & \cellcolor{Gray}        \\ 
\hline
iOS-C & \cellcolor{Gray}         &    No      &  \cellcolor{Gray}        &     Yes       & \cellcolor{Gray}        &   Yes      \\ 
\hline
iOS-P &   Yes       & \cellcolor{Gray}         &    Yes      &  \cellcolor{Gray}         &   Yes      &  \cellcolor{Gray}       \\ 
\hline
\end{tabular}
\end{table}
%

@@

%
\subsection{Power constraints}
\label{sec:power}
%
Due to indefinite foreground services and continuous discovery and
advertisement, this is primarily an issue on Android. Our approach
here is two-pronged. Firstly, via extensive experiments on a large
number of devices, we fine-tune the algorithms governing the intervals
at which discovery and advertisements occur. In a nutshell, receiving
new information during a period causes MNA to be more aggressive in
discovery and advertisement. Similarly, lack of new information for a
period makes the device less aggressive. Secondly, we allow the
central service to broadcast``wake up'' messages ahead of an
anticipated severe weather event.  When devices receive such messages,
they schedule themselves to remain aggressive during the specified
window of time.  Outside of this window, the devices can afford to have
long sleep cycles and conserve power. With these techniques, our
testing shows less than 2\% battery consumption per hour on most
device models.
%
\subsection{Testing p2p networks}
\label{sec:texting}
%
We developed novel test automation tools and processes that control
multiple devices from a single test station and follow prescribed
steps to generate, send, and receives messages to play out a test
scenario.  The framework allows automated analysis of the observations
to determine the test result. This capability was instrumental in
uncovering bugs at a fast pace with tens of physical devices employed
for automated testing.  Further, the automation framework is general
and can be expressly applied to test other apps in this fashion.

Figure~\ref{fig:test_arch} shows the test automation architecture. A
tester specifies high-level test cases and expected results in a
platform and device-agnostic fashion. Mobile devices run a UI proxy
app for executing platform and device-specific UI commands while a
central laptop terminal acts as the overall orchestrator. The
orchestrator translates high-level test cases and distributes them to
devices. Since the orchestrator needs wired access to the devices and
the devices cannot be programmed to move, testing of range and
mobility patterns needs to done manually, which is a known limitation.

A key innovation here is the platform-independent language for
specifying high-level test cases and a small set of primitive commands
it translates to. We describe here the list of primitive commands and
have made available a screen recording across seven Android devices
demonstrates this powerful capability \cite{mna-test-demo}.
\begin{itemize}
\item \textsf{syncAll} -- Wait for all devices to reach this point
\item \textsf{object.find} -- Find a UI object by name or position
\item \textsf{object.click} -- Click a UI object
\item \textsf{sleep} -- Wait for the specified duration
\item \textsf{setupSystemAlert} -- Respond to a system popup
\item \textsf{text.set} -- Highlight a text input box
\item \textsf{text.write} -- Write text to the text box
\item \textsf{swipe} -- Swipe to the left/right or up/down
\end{itemize}

@@

%
\subsection{Trust in information}
\label{sec:trust}
%
In c2p2p, since the central service originates all messages, it can
digitally sign them. The mobile application is distributed
with the corresponding public key so that digital signatures from the
service can be verified.  If a message fails such a verification, it
is discarded and not forwarded any further.

Peers cannot originate messages however as then a malicious peer can
join the network, exchange keys with other peers, start generating,
signing, and spreading fake information. This has been a longstanding
problem for open decentralized systems. Requiring all peers to
register their identity with a certificate authority and reporting the
originators of fake information may discourage malicious behavior.
Clearly, this is even harder to solve than the spread of fake
information over the Internet because ground truth is hard to find and
malicious peers can report genuine sources of information as fake.
%
\section{Algorithms}
\label{sec:algo}
%
\subsection{WiFi DNS}
\label{sec:wifi}
%
DNS service discovery is a widely supported protocol
\cite{cheshire-dns-sd-2013}. On Android, WiFi DNS improvises on this
standard such that communication happens without making network
connections. This is achieved by splitting the message payloads into
small enough chunks, stuffing them into service advertisements as txt
records, and broadcasting them over multiple advertisements. Since
there are no connections being made, the messages can be exchanged
without prompting the user to grant permissions for each
connection. Since no peer connections are needed and advertisements
and discovery are broadcast, WiFi DNS is a broadcast
protocol. Although the standard states that as much as 65KB of txt
records can be added to advertisements, in practice most devices
support only up to 750 bytes. On iOS, Bonjour protocol is built on DNS
service discovery but is power-hungry and does not inter-operate with
WiFi DNS on Android. Algorithm~\ref{algo:dns} outlines the peer
process for WiFi DNS. Power conservation via time intervals is
achieved by manipulating timers $T_{ad}$ and $T_{disc}$ based on
traffic. $\mathbb{M}$ is the set of all unexpired messages.

Messages from multiple applications are forwarded on the same network,
but a peer can validate signatures only for the messages targeting an
app it hosts. This is because peers do not possess the public keys
corresponding to the central services of other apps. Hence, peers
simply forward messages not targeting one of their apps. Key benefit
of this design is that the entire network of peers is available to all
apps, making c2p2p scenarios viable even when an individual app does
not have a sufficient density of users.  However, this also allows a
malicious jammer to generate traffic for a fake app, forcing the peers
to forward it until the messages expire.  Although the fake
information will never be delivered to the apps, this affects
throughput. A possible solution would be to distribute a registry of
valid $a_{id}$ to all peers across apps so that they can ignore
messages with fake $a_{id}$ and not forward them.

\begin{algorithm}[htb!]
\label{algo:dns}
\DontPrintSemicolon % Some LaTeX compilers require you to use \dontprintsemicolon instead
\SetAlgoLined
\KwIn{Peer id $p_{id}$, App id $a_{id}$}
\KwIn{max advert/discovery intervals $max_{ad}$, $max_{disc}$}
\KwIn{min advert/discovery intervals $min_{ad}$, $min_{disc}$}
 $S_{name} \gets$ \textsf{'MNA'}\;
 $\mathbb{M} \gets \varnothing$\;
 $T_{ad} \gets min_{ad}$\;
 $T_{disc} \gets min_{disc}$\;
 $New \gets 0$\;
 \ForEach{timer expiry $T_{ad}$}{
   $\mathbb{M} \gets \mathbb{M} -$ expired messages\;
   $\mathbb{M}_{sort} \gets$ $\mathbb{M}$ sorted by least advertised first\;
   $B \gets$ $\mathbb{M}_{sort}$ chunked by MTU size\;
   dnsAdvertise ($S_{name}$, $p_{id}$, $B$)\;
   \If{$New = 0$}{
     $T_{ad} \gets min(T_{ad} \times 2, max_{ad})$\;
   }
 }
 \ForEach{timer expiry $T_{disc}$}{
   dnsDiscover ($S_{name}$, $p_{id}$)\;
   \If{$New = 0$}{
     $T_{disc} \gets min(T_{disc} \times 2, max_{disc})$\;
   }
   $New \gets 0$\;
 }
 \ForEach{discovered service callback $S_{disc}$}{
   $B \gets S_{disc}.B$\;
   $M_{part} \gets msg(B)$\;
   $M \gets M + M_{part}$\; 
   \If{$M_{part}$ is not a duplicate}{
        $T_{ad} \gets max(\frac{T_{ad}}{2}, min_{ad})$\;
        $T_{disc} \gets max(\frac{T_{disc}}{2}, min_{disc})$\;
        $New \gets 1$\;
   }
   \If{$M_{part}$ was the last chunk of $M$}{
     \If{$M$ is unexpired}{
         \If{$M.app = a_{id}$ $\wedge$ $M$ signature is valid}{
             Notify app with $M$\;
         }
         $\mathbb{M} \gets \mathbb{M} \cup M$\;
     }
   }
 }
 \tcc{For testing only}
 \ForEach{app request to send $M$}{
   \If{$M$ has not expired}{
     $\mathbb{M} \gets \mathbb{M} \cup M$\;
   }
 }
 \caption{WiFi DNS peer algorithm}
\end{algorithm}
%
\subsection{BTLE}
\label{sec:btle}
%

The only protocol that stays reliably active in background without
violating developer guidelines is BTLE. Unlike Android, iOS
applications do not need to discover and advertise continually. Once a
BTLE Peripheral (role that has information) is advertised and a
Central (role that receives information) requests discovery, the app
is allowed to go to sleep. When a matching Central or a Peripheral is
found, iOS wakes up the application to handle such an event, even when
the application is in the background. MNA builds on this and leverages
asynchronous dispatch queues to turn each peer into a Central as well
as a Peripheral simultaneously. This architecture allows bidirectional
data transfers without concurrency issues.

Algorithm~\ref{algo:btle} outlines the peer process for BTLE. The
biggest difference with WiFi DNS is that discovery and advertisements
happen only once per app launch. Secondly, the entire peer process is
event-driven, so the app can sleep when there are no events to
process. The list of forwarders for each message $M.fwdList$ is key in
controlling duplicate transmissions without overhead messages. In
practice, limiting number of Peripherals a Central connects to enables
reliable communications, specified by $max_p$.
